Modular analysis of gene expression data with R

### 2009-12-11 isanote/modular.tex
More isa application note changes.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{\subfig{(A)} Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package. See the text for details.
  Subfigures \subfig{B} and \subfig{C} were generated
  using the acute lymphoblastic leukemia data set, see
  \citep{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(B)} Heatmap for a single module, showing correlated
  expression for the genes and samples. The red lines are the gene and
  sample scores.
  \subfig{(C)} A module tree. Each rectangle is a bicluster; see the
  definition of the edges in the text. The modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the bicluster. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
}
\label{fig:workflow}
\end{figure*}

@@

% Batch correction
\emph{Batch correction.}
To study the global organization of a transcription program including
many aspects of transcriptional regulation one often combines several
microarray experiments into a single dataset. In such a case
additional data normalization is vital to reduce the bias due to
constituent datasets. Several methods address this challenge,
see e.g.~\citep{johnson07} for an algorithm that has a GNU R
implementation. 
% (S: Do we have a standard flag to call their software?)

% Filtering
\emph{Gene filtering.}
Genes that have very low expression levels in all samples, carry little if
any information and may reflect ineffective array probes, etc. Since these
genes are likely to contribute only noise to the analysis, we suggest
removing them before running the ISA iteration.
% (S: Describe how this is done, and if it requires setting a threshold!)

% Normalization.
\emph{ISA normalization.} (Step 1 on Fig.~\ref{fig:workflow}.)
In each iteration of the ISA thresholded weighted sums of expression
levels over either genes or samples are computed. Since different genes
typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
calculated for each gene across all samples, the other for each sample
across all genes.

% Random and smart seeding.
\emph{Random and smart seeding, ISA iteration} (Step 2.)
The ISA iteration is typically performed to a
large number of seeds. In the unsupervised approach these seeds are chosen
randomly to sample uniformly the immense search space. We also implemented a
semi-supervised method, to which we refer as ``smart seeding'', where the
seeds a biased to start with certain sets of genes or samples based on
prior knowledge.As mentioned previously, ISA can be performed with random or smart
seeds, based on the application.

% Merging the modules.
\emph{Merging the modules.} (Step 3.)
It is possible that several seeds converge to the same, or very similar
biclusters. This step eliminates such duplicates.

% The robustness measure and filtering the modules.
\emph{Robustness of biclusters.} (Step 3.)
To access the significance of a bicluster, we designed a robustness
measure, that can be used to filter out spurious modules. This is done
based on performing ISA on the scrambled input data.

% Module trees.
\emph{Module trees.}
The ISA works with two stringency threshold parameters, the gene
threshold and the sample threshold. ISA modules can be organized into
a directed graph, a module tree, in which there is an edge from module
$A$ to module $B$, if the ISA converges to module $B$ from module $A$,
with the same threshold parameters  that were used to find module
$B$. A module tree provides a hierarchical modular description of a
data set.

@@

Both the \Rpackage{isa2} and \Rpackage{eisa} packages support two
workflows. The \emph{simple workflow} involves a single R function
call and runs all ISA steps (1-3 in Fig.~\ref{fig:workflow}.) with
their default parameters.

The \emph{detailed workflow} means running every step of the modular
analysis separately, possibly with non-default parameters. This allows
the users to tailor the ISA completely according their needs.

@@

The \Rpackage{eisa} package implements a set of visualization
techniques for biclusters, one of which is specific to ISA (see
Fig.~\ref{fig:workflow} for examples).

% The \Rpackage{ExpressionView} R package, discussed in detail in
% another application note, implements an interactive tool
% for visualizing ISA (or other) overlapping biclusters.

% \subsubsection{Connection to other software}

@@

\paragraph{Funding\textcolon} The authors are greatful to the Swiss
Institute of Bioinformatics, the Swiss National Science Foundation
(3100AO-116323/1), and the European Framework Project 6 (through
the EuroDia and AnEuploidy projects).
\paragraph{Conflict of interest\textcolon} None declared.

>>>>

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{\subfig{(A)} Work flow of a typical modular analysis with the
  \Rpackage{eisa} package. See text for details.
  Subfigures \subfig{B} and \subfig{C} were generated
  using the acute lymphoblastic leukemia data set, see
  \citep{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(B)} Heatmap for a single module, showing coherent
  expression of the genes across the samples. The red lines are the gene and
  sample scores.
  \subfig{(C)} Module tree. Each rectangle is a module; see the
  definition of the edges in the text. Modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the module. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
}
\label{fig:workflow}
\end{figure*}

@@

% Batch correction
\emph{Batch correction.}
To study the global organization of a transcription program including
many aspects of transcriptional regulation one often combines several
microarray experiments into a single dataset. In such a case
additional data normalization is crucial to reduce the bias due to
constituent datasets. Several methods address this challenge,
see e.g.~\citep{johnson07} for an algorithm that has a GNU R
implementation.

% Filtering
\emph{Gene filtering.}
Genes that have very low expression levels in all samples, carry little if
any information and may reflect ineffective array probes, etc. Since these
genes are likely to contribute only noise to the
analysis~\citep{hackstadt09}, we suggest removing them before running
the module identification of the ISA.

% Normalization.
\emph{ISA normalization.} (Step 1 on Fig.~\ref{fig:workflow}.)
In each iteration the ISA computes thresholded weighted sums of expression
levels over either genes or samples. Since different genes 
typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
calculated for each gene across all samples, the other for each sample
across all genes.

% Random and smart seeding.
\emph{Random and smart seeding, ISA iteration} (Step 2.)
The iterative procedure of the module identification is typically applied to a
large number of seeds. In the unsupervised approach these seeds are chosen
randomly to sample uniformly the immense search space. We also implemented a
semi-supervised method, to which we refer as ``smart seeding'', where the
seeds are biased to start with certain sets of genes or samples based on
prior knowledge. The ISA can be performed with random or smart
seeds, depending on the application.

% Merging the modules.
\emph{Merging and filtering the modules.} (Step 3.)
It is possible that several seeds converge to the same, or very similar
biclusters. This step eliminates such duplicates.
To access the significance of a module, we designed a robustness
measure, that can be used to filter out spurious modules. This is done
by applying the ISA to scrambled input data in order to obtain a
reference (null) distribution for the significance scores.

% Module trees.
\emph{Module trees.}
The ISA works with two stringency threshold parameters, the gene
threshold and the sample threshold. ISA modules can be organized into
a directed graph, to which we refer as ``module tree''. An edge from module
$A$ to module $B$ indicates that the ISA converges to module $B$ from module $A$,
with the same threshold parameters that were used to find module
$B$. A module tree provides a hierarchical modular description of a
dataset.

@@

Both the \Rpackage{isa2} and \Rpackage{eisa} packages support two
workflows. The \emph{simple workflow} involves a single R function
call and runs all ISA steps (1-3 in Fig.~\ref{fig:workflow}) with
their default parameters.

In the \emph{detailed workflow} every step of the modular analysis 
is executed separately, possibly with non-default parameters. This allows
the users to tailor the ISA according their needs.

@@

The \Rpackage{eisa} package implements a set of visualization
techniques for modules (see Fig.~\ref{fig:workflow} for examples).

@@

\paragraph{Funding\textcolon} The authors are grateful to the Swiss
Institute of Bioinformatics, the Swiss National Science Foundation
(3100AO-116323/1), and the European Framework Project 6 (through
the EuroDia and AnEuploidy projects).
\paragraph{Conflict of interest\textcolon} None declared.
